---
title: "Brief overview in R"
author: "Nima Rafati"
date: "2024-04-21"
output: 
  html_document: 
    fig_width: 5
    fig_height: 4
editor_options: 
  chunk_output_type: inline
---


```{bash}
echo "This is a test"

```

 
# Starting with data
  - Description of the data
  - draw a table with index
  - data_raw
  - c(), round()
```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```
When you load tidyverse:
- dplyr: collection of tools t easily manipulate the data 
- tidyr: reshaping the data for plotting
- ggplot2: for advanced plotting
- tibble: modern dataframe


**[`dplyr` cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)**      
**[`tidyr` cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-import.pdf)**  


```{r, eval=FALSE, purl=TRUE}
download.file(
  url = "https://nbisweden.github.io/module-r-intro-dm-practices/data/Hawks.csv",
  destfile = "data_raw/Hawks.csv"
)
```


read_csv from `tidyverse` package == read.csv from base R.  

```{r, echo = FALSE, eval = TRUE, purl = FALSE}
# silently read in CSV file with data
hawks <- read_csv("data_raw/Hawks.csv") #, col_names = T)
```

When you execute `read_csv` on a data file, it looks through the first 1000 rows 
of each column and guesses its data type. Use the function `spec()` to view the
full column specification:

## Dataframes
columns = vector (datatype)

## Inspecting the dataframes

```{r, eval=TRUE,  purl=FALSE}
view(hawks) 
hawks
head(hawks)
tail(hawks)
str(hawks)#structure
spec(hawks)
summary(hawks)

nrow(hawks)
ncol(hawks)
dim(hawks)
```

```{r}

hawks$CaptureTimeptureTime[!is.character(hawks$CaptureTime)]
```

>>>>>>>>>>Challenge 2.1 

## Indexing and subsetting dataframes 

```{r, purl=FALSE}
# first element in the first column of the data frame
hawks[1, 1]   
# first element in the 6th column
hawks[1, 6]   
# first column of the data frame
hawks[, 1]    
# first column of the data frame
hawks[1]      
# first three rows of the 6th column
hawks[1:3, 6] 
# the 3rd row of the data frame
hawks[3, ]    
# equivalent to head_hawks <- head(hawks)
head_hawks <- hawks[1:6, ] 

# select columns by name
hawks$Month[1:6]
hawks[,'Month']
``` 

```{r}
hawks[1,c('Year', "ReleaseTime", 'Sex')]
sum(!is.na(hawks$Sex))

hawks$Sex[!is.na(hawks$Sex[1:6])]

hawks[1:6,][is.na(hawks$Sex[1:6]),]

hawks_known_sex <- hawks[!is.na(hawks$Sex),]
dim(hawks_known_sex)
```

## Factor
Categorical data
```{r}
levels(hawks$Sex)
hawks$Sex <- factor(x = hawks$Sex, levels = c('M', 'F'), labels = c('Male', 'Female'))
levels(hawks$Sex)
summary(hawks$Sex)
```

## Plotting
```{r fig.height = 6, fig.width = 6, fig.align = "center", purl = FALSE}
plot(hawks$Sex)
```

```{r}
sex <- hawks$Sex
levels(sex)
sex <- addNA(sex)
levels(sex)
levels(sex)[3] <- 'Unknown'
'Unknown1' -> levels(sex)[3] 
```

```{r barplot-1, fig.height = 6, fig.width = 6, fig.align = "center", purl = FALSE}
plot(sex)
```


## Bonus

```{r}
x <- 1:10
exp_x <- exp(x)
plot(exp_x, type = 'l', xlab = 'time', ylab = 'progress')
lines(x * 500, col = 'red')
```

# Data manipulation
```{r}
hawks <- read_csv("data_raw/Hawks.csv") #, col_names = T)
```

## select columns

```{r}
select(hawks, Species, Sex, Weight)
select(hawks, -BandNumber, -Culmen)
```


```{r, purl = FALSE}
filter(hawks, Sex == "F")
```

We can also filter rows that do not contain missing data in some columns:

```{r, purl = FALSE}
filter(hawks, !is.na(Sex) & !is.na(Weight))
```

This will return all rows that have a value in both the `Sex` column and the
`Weight` column. In **`Tidyverse`**, there is also a special functions `drop_na` 
that can be used to filter out rows with missing data:

```{r, purl = FALSE}
drop_na(hawks, Sex, Weight)
```


## Pipes (`%>%`)

Ctrl + Shift + M
Cmd + Shift + M


x %>% f ==> f(x)
x %>% f(y) ==> f(x,y)


```{r}
hawks[1:5,]
head(hawks, n = 5)
hawks %>% head(n = 5)
hawks %>% head(n=5) %>% tail(n=1)

hawks %>%
  filter(Sex == "F" | Weight > 500) %>%
  select(Species, Sex, Weight)
```

>>>>>>>> Challenge 3.1

## Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

```{r, purl = FALSE}
colnames(hawks)
hawks %>%
  mutate(Weight_kg = Weight / 1000)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl = FALSE}
hawks %>%
  mutate(Weight_kg = Weight / 1000,
         Weight_lb = Weight_kg * 2.2)
```

The first few rows of the output are full of `NA`s, so if we wanted to remove
those we could insert a `filter()` in the chain:

```{r, purl = FALSE}
hawks %>%
  filter(!is.na(Weight)) %>%
  mutate(Weight_kg = Weight / 1000) %>%
  head()
```


>>>>>>>> Challenge 3.2

## Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of
the `group_by()` function.

### `summarize`
```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Sex) %>%
  summarize(mean = mean(Weight, na.rm = TRUE))
```

You can also group by multiple columns:
```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Species, Sex) %>%
  summarize(mean = mean(Weight, na.rm = TRUE))
```

More metrics 
```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Species, Sex) %>%
  summarize(mean = mean(Weight, na.rm = TRUE),
            min = min(Weight, na.rm = TRUE))
```

It is sometimes useful to rearrange the result of a query to inspect the values.
For instance, we can sort on `min` to put the lowest numbers first:

```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Species, Sex) %>%
  summarize(mean = mean(Weight, na.rm = TRUE),
            min = min(Weight, na.rm = TRUE)) %>% 
  arrange(min)
```

```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Species, Sex) %>%
  summarize(mean = mean(Weight, na.rm = TRUE),
            min = min(Weight, na.rm = TRUE)) %>% 
  arrange(desc(min))
```

### Counting
When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each Sex, we would do:

```{r, purl = FALSE}
hawks %>%
  count(Sex) 
```

The `count()` function is shorthand for something we've already seen: grouping
by a variable, and summarizing it by counting the number of observations in that
group. In other words, `hawks %>% count(Sex)` is equivalent to:  

```{r, message = FALSE, purl = FALSE}
hawks %>%
  group_by(Sex) %>%
  summarize(n = n())
```

We can also combine `count()` with other functions such as `filter()`. Here
we will count the number of each species with weights above 800 g.

```{r, purl = FALSE}
hawks %>%
  filter(Weight > 500) %>%
  count(Species)
```

The example above shows the use of `count()` to count the number of
rows/observations for *one* factor (i.e., `Species`). If we wanted to
count *combination of factors*, such as `Species` and `Sex`, we would
specify the first and the second factor as the arguments of `count()`:

```{r purl = FALSE}
hawks %>%
  filter(Weight > 500) %>%
  count(Species, Sex)
```

With the above code, we can proceed with `arrange()` to sort the table according
to a number of criteria so that we have a better comparison. For instance, we
might want to arrange the table above in (i) an alphabetical order of the levels
of the sex and (ii) in descending order of the count:

```{r purl = FALSE}
hawks %>%
  filter(Weight > 500) %>%
  count(Species, Sex) %>% 
  arrange(Sex, desc(n))
```


>>>>>>>> Challenge 3.4



# Data Visualization with ggplot2

- Data: Input data
- Aesthetic: Visual characteristics of the geometry. Size, Color, Shape etc
- Geom: A geometry representing data. Points, Lines etc
- Scale: How visual characteristics are converted to display values
- Statistics: Statistical transformations. Counts, Means etc
- Coordinates: Numeric system to determine position of geometry. Cartesian, Polar etc
- Facets: Split data into subsets
```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>() ......
```

- Stepwise (iteratively)

## points
```{r}
colnames(hawks)
ggplot(mapping = aes(x = Weight, y = Wing),
       data = hawks)
# size, color, shape,

ggplot(data = hawks, 
       mapping = aes(x = Weight, y = Wing)) + 
  geom_point()
```

## coloring
```{r}
ggplot(data = hawks, 
       mapping = aes(x = Weight, y = Wing, dill)) + 
  geom_point( color = 'blue')

ggplot(data = hawks, 
       mapping = aes(x = Weight, y = Wing)) + 
  geom_point( color = 'blue', alpha = 0.5)

ggplot(data = hawks, 
       mapping = aes(x = Weight, y = Wing)) + 
  geom_point( aes(color = Species), alpha = 0.5)
```


>>>>>>>>> Challenge 4.2

## Boxplot

```{r}
ggplot(data = hawks, mapping = aes(x = Species, y = Wing)) +
    geom_boxplot()
```


```{r boxplot-with-points, purl=FALSE}
ggplot(data = hawks, mapping = aes(x = Species, y = Wing)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5, color = "tomato")
```



>>>>>>> Challenge 4.3

```{r}

```

## Barplot

```{r}
# filter
hawks %>% head( n = 5) %>% 
  filter(Species == 'CH' | Species == 'RT')

# count table
hawks %>% head( n = 5) %>% 
  filter(Species == 'CH' | Species == 'RT') %>%
  count(Year, Species)

```


```{r}
sex_counts <- hawks %>% count(Sex)
sex_counts

ggplot(data = sex_counts, aes(x = Sex, y = n)) + geom_bar(stat = 'identity')

# color each bar by sex ==> fill

```


## Time series data

```{r, purl = FALSE}
yearly_sp_counts <- hawks %>%
  filter(Species  == "CH" | Species == "SS") %>% 
  count(Year, Species)
```

```{r timeseries1, purl = FALSE}
ggplot(data = yearly_sp_counts, aes(x = Year, y = n)) +
  geom_line()

# What is the problem?
# pipe to ggplot
```

```{r timeseries2, purl=FALSE}
ggplot(data = yearly_sp_counts, aes(x = Year, y = n, group = Species)) +
  geom_line()
```



```{r timeseries3, purl=FALSE}
ggplot(data = yearly_sp_counts, aes(x = Year, y = n, color = Species)) +
  geom_line()
```

## Faceting
Split the data 
```{r facet1, purl = FALSE}
ggplot(data = yearly_sp_counts, aes(x = Year, y = n)) +
    geom_line() +
  facet_wrap(facets = vars(Species))
```

Now we would like to split the line in each plot by the sex of each individual 
measured. To do that we need to make counts in the data frame grouped by `Year`, 
`Species`, and `Sex`. We will also drop rows with `NA` in any of these three columns.


```{r, purl = FALSE}
yearly_sp_sex_counts <- hawks %>%
  filter(Species  == "CH" | Species == "SS") %>% 
  drop_na(Year, Species, Sex) %>% 
  count(Year, Species, Sex)
```

We can now make the faceted plot by splitting further by sex using `color` 
(within a single plot):

```{r facet2, purl = FALSE}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
    geom_line() +
  facet_wrap(facets = vars(Species))
```
You can also organise the panels only by rows (or only by columns):

```{r facet-in-rows, purl = FALSE}
# One row, facet by column
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
    geom_line() +
    facet_grid(rows = vars(Species))
```

It is also possible to create more advanced layouts using the `facet_grid()`
function. We can for example facet both by species and sex:

```{r facet3, purl = FALSE}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex, group = Sex)) +
    geom_line() +
  facet_grid(rows = vars(Sex), cols = vars(Species))
```
## Themes

```{r}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
    geom_line() +
    facet_wrap(facets = vars(Species)) +
    theme_bw() # theme_void, theme_minimal, theme_light
```

>>>>>>> Challenge 4.4

```{r}

```

## Customisation 

- Cheats heet

```{r}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
  geom_line() +
  facet_wrap(vars(Species)) +
  labs(title = "Captured individuals through time",
     x = "Year of capture",
     y = "Number of individuals") +
  theme_bw()
```


```{r}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
  geom_line() +
  facet_wrap(vars(Species)) +
  labs(title = "Captured individuals through time",
     x = "Year of capture",
     y = "Number of individuals") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```


Increase the font size of the axis

```{r}
ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
  geom_line() +
  facet_wrap(vars(Species)) +
  labs(title = "Captured individuals through time",
     x = "Year of capture",
     y = "Number of individuals") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), text = element_text(size = 16))
```

## Questions if not Challenge 4.5

## Bonus: Arranging plots

- `patchwork`
- `cowplot`
```{r}
library(patchwork)
p1 <- ggplot(data = hawks, mapping = aes(x = Species, y = Wing)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5, color = "tomato")
p2 <- ggplot(data = yearly_sp_sex_counts, aes(x = Year, y = n, color = Sex)) +
  geom_line() +
  facet_wrap(vars(Species)) +
  labs(title = "Captured individuals through time",
     x = "Year of capture",
     y = "Number of individuals") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 16))
p1 | p2
```

```{r}
library(cowplot)
p3 <- cowplot::plot_grid(p1,p2,ncol = 1)

ggsave('summary_plots.png', p3, width = 25, height = 10)
```

## Bonus 

```{r}
# Load ggplot2 library
library(ggplot2)

# Assume the dataset is already loaded and named 'Hawks'
# If you need to load data from a file:
# Hawks <- read.csv("path_to_hawks_data.csv")

# Plotting weight versus wing span with density color
ggplot(data = hawks, aes(x = Weight, y = Wing)) +
  geom_point(aes(color = ..density..), stat = "density2d", size = 3) +  # Using density2d to color points by density
  scale_color_gradient(low = "yellow", high = "red") +  # Color gradient from yellow (low density) to red (high density)
  labs(title = "Weight vs. Wing Span in Hawks",
       x = "Wing Span (cm)",
       y = "Weight (g)",
       color = "Point Density") +
  theme_minimal()

```



```{r}
# Library
library(ggplot2)

# Dummy data
x <- LETTERS[1:20]
y <- paste0("var", seq(1,20))
data <- expand.grid(X=x, Y=y)
data$Z <- runif(400, 0, 5)
 
# Heatmap 
ggplot(data, aes(X, Y, fill= Z)) + 
  geom_tile()


# Reshape the data into a matrix
mat <- matrix(data$Z, ncol = length(unique(data$X)), byrow = TRUE)

# Create heatmap
heatmap(mat, Rowv = NA, Colv = NA, xlab = "X", ylab = "Y", col = heat.colors(20))

```












